\textbf{NBA Party}

### 2017-07-06 cardinal_analytx_presentation/nightlife_and_sports.tex
Changed frametitle's from Camel Case to proper punctuation. Changed ordering of 'Whats our Solution'.

\author{Andreas Santucci} \institute{Stanford Teaching Fellow \\ Computational Math, Data Science} \title{Effects of nightlife on sports} \subtitle{Cardinal Analytx Interview} % \titlegraphic{\includegraphics[width=2cm]{ctanlion}}
 \begin{document}  \begin{frame} \maketitle \end{frame}

@@

    \vspace{12pt}It's not unreasonable to think that what happens off the court affects game-outcomes, but!
    \begin{itemize}       \item Can we even measure this?
      \item Are there systematic tendencies?     \end{itemize}
  \end{block} \end{frame}

@@

      \vspace{12pt}After all, we can't even observe if players party, let alone when.

@@

\begin{frame}
  \frametitle{Previous Research - NBA player shooting average}     
  NBA Players under-perform proportion of shots made on Sundays.\footnote{Chase, ('15).}
    \begin{itemize}       
    \item 1,092 of 2,035 NBA players who took $\geq 1$ shot on Sundays under-perform 
      relative to their own season-long average; $p=0.001$.
      \item Effect size is not considered.
      \item Density of support is questionable.\footnote{What if we miss the only shot we took on Sunday, is that relevant?}     \end{itemize}
\end{frame}

\begin{frame}   \frametitle{Prev. Research - Likelihood of meeting the spread in NBA}
  Teams under-perform bookmakers expectations on Sunday afternoons when on the road.\footnote{Ezekowitz, ('14).}
\vspace{12pt}
    \begin{itemize}       
    \item Only 129 out of 284 games did the away team meet the spread, 
      using 2006-14 data; $p=0.1089$.
    \item Not quite significant at the 90\% confidence level.     
    \item If we bet \$100 on each Sunday game, we'd only profit \$1,310 over 8 years.\footnote{$(284 - 129) \times 100 + 129 \times (-110) = 1,310$.} Not exciting.
    \end{itemize}
  \end{frame}

\begin{frame}   
  \frametitle{What is our solution?}
  \begin{block}{We can't observe whether players party.}
    Instead, we will consider a quasi-experiment. Motivating questions:
    \begin{itemize}       \item Do teams play worse the day after visiting a ``party city''?
      \item How do we even determine whether a city has nightlife?    
      \item What does it mean to ``play worse''? \end{itemize}
  \end{block}

@@

    Conditions on every conceivable element of the game:
    \begin{itemize}     \item strengths and weaknesses against particular opponents,
    \item momentum,
    \item injuries, and
    \item even possibly a weather forecast.     \end{itemize}

@@

\begin{frame}   \frametitle{Quasi Experiment}
  Consider games played back-to-back within 24 hours.

@@

\begin{frame}   \frametitle{NBA Data: 2010-11 through 2016-17 season}
  \begin{itemize}     \item Each team plays 82 games per season.
    \item 30 teams in the NBA.
    \item Total of $82 \times 30 / 2 = 1,230$ games per season.  
  \end{itemize}

@@

\begin{frame}   \frametitle{MLB Data: 2011 through 2017 season}
  \begin{itemize}     \item Each team plays 162 games per season.
    \item 30 teams in the MLB.
    \item Total of $162 \times 30 / 2 = 2,430$ games per season.   \end{itemize}

@@

\begin{frame}   \frametitle{Bureau of Labor Statistics Data: 2010-2016}
  We look toward BLS data which describes the number of business establishments 
  (of a very particular type) in each metropolitan statistical area
  for each quarter of each year.

@@

\begin{frame}   
  \frametitle{Possible Confounders}
  What might influence game-outcomes apart from player-ability?

@@

\begin{frame}   \frametitle{Feature Creation}
  We add features which describe
  \vspace{12pt}  
  \begin{itemize}     
    \item Number of days since the last game (i.e. rest-time),
    \item Lagged changes in possession,\footnote{A crude measure of fatigue incurred from last game.}
    \item Lagged game location, and
    \item Travel Distance.   
  \end{itemize}
  \vspace{12pt}We seek to construct a model which accounts for travel-fatigue and jet-lag.
  For this, we need a little bit more. \end{frame}

\begin{frame}   \frametitle{Accounting for Jetlag}
  Prolonged air travel negatively effects sporting performance.\footnote{Lee and Galvez ('12).}

@@

\begin{frame}   
  \frametitle{Travel Distance Comparison}
  \centering \includegraphics[scale=0.25]{../writing/travel_dist_density_by_party} 
\end{frame}

\begin{frame}   \frametitle{Continuous Measure}
  \begin{block}{A-list celebrities tend to party with other celebrities.}     Idea: proxy nightlife using number of musicians in previous location

@@

\begin{frame}   
  \frametitle{Continuous Measure of Party (NBA)}
  \centering \includegraphics[scale=0.3]{../writing/Party_by_Team_Location} 
\end{frame}

@@

\begin{frame}   \frametitle{Placebo Test (discrete indicator): what if players rest?}
  \centering
  \tiny{
\begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{1}{c}{\textit{Dependent variable:}} \\  \cline{2-2}  \\[-1.8ex] & Meet the Spread (NBA) \\  \hline \\[-1.8ex]   Party placebo & 0.072 \\    & (0.082) \\    & \\   Lag changes in posession & 0.00001 \\    & (0.002) \\    & \\   Logged travel distance & $-$0.005 \\    & (0.026) \\    & \\   East-west travel direction & $-$0.099 \\    & (0.231) \\    & \\   Number hours rest time & 0.001$^{*}$ \\    & (0.001) \\    & \\   Time of game during day & 0.001 \\    & (0.013) \\    & \\   Home team effect & $-$0.009 \\    & (0.043) \\    & \\   Logged travel distance * east-west & 0.015 \\    & (0.036) \\    & \\   Constant & $-$0.069 \\    & (0.378) \\    & \\  \hline \\[-1.8ex]  Observations & 9,823 \\  Log Likelihood & $-$6,806.701 \\  Akaike Inf. Crit. & 13,631.400 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}  } \end{frame}

@@

\begin{frame}   
  \frametitle{Continuous Measure of Party (MLB)}
  \centering \includegraphics[scale=0.3]{../writing/Party_by_Team_Location_mlb}   
\end{frame}

@@

\begin{frame}   \frametitle{Returning to NBA: Player Performance}
  We additionally look toward actual performance metrics, instead of relying on point-spread data.

@@

\begin{frame}   \frametitle{Future Research - NBA}
  Within the NBA
  \begin{itemize}     
    \item Are teams more likely to party after a win or a       loss?
    \item Do teams who aren't going to make the playoffs party more?
    \item Are there demographic differences which influence propensity to party? E.g. are younger, unmarried players more susceptible to our treatment effect?
    \item We could also examine how shots taken vs. shots scored are affected for star players.
  \end{itemize}
  \vspace{12pt}  A preliminary and ad-hoc analysis shows that injury risk increases as a function of party.\footnote{We look for instances where players are put on the bench and stop playing (at least several) games as a function of our nightlife indicator.}
  This might inform coaches which players to bench and when.

@@

\begin{frame}   \frametitle{Future Research (MLB and General)}
  \begin{itemize}     
  \item Within the MLB, we can look at statistics such as batting averages or general fielding statistics. 
    \item Possibly prolonged effects; recall our magic 24 hour number.
    \item European soccer leagues.   \end{itemize} \end{frame}

@@

\begin{frame}   \frametitle{Why \# Drinking Establishments or Population won't work}
  \includegraphics[scale=0.3]{../writing/why_drinks_or_population_dont_work} \end{frame}

>>>>

\author{Andreas Santucci} \institute{Stanford Teaching Fellow \\ Computational Math, Data Science} \title{Effects of nightlife on sports} \subtitle{``Hangover Effects''} % \titlegraphic{\includegraphics[width=2cm]{ctanlion}}
 \begin{document}  \begin{frame} \maketitle \end{frame}

@@

    \vspace{12pt}It's not unreasonable to think that what happens off the court affects game-outcomes, but!
    \begin{itemize}       
      \item Can we even measure a hangover effect?
      \item Are there systematic tendencies?
      \end{itemize}
  \end{block} 
\end{frame}

% What is our solution?

\begin{frame}   
  \frametitle{Whether players actually party is a latent variable}
  \begin{block}{We can't observe whether players party.}
    Instead, we will consider a quasi-experiment where players are exposed to nightlife. 
    Motivating questions:
    \vspace{12pt} 
    \begin{itemize}       \item Do teams play worse the day after visiting a ``party city''?
      \item How do we even determine whether a city has nightlife?    
      \item What does it mean to ``play worse''? \end{itemize}
  \end{block}

@@

      \vspace{12pt}After all, we can't reliably observe if players party, let alone when.

@@

\begin{frame}
  \frametitle{Previous research - NBA player shooting average}     
  NBA Players under-perform proportion of shots made on Sundays.\footnote{Chase, ('15).}
    \begin{itemize}       
    \item 1,092 of 2,035 NBA players who took $\geq 1$ shot on Sundays under-perform 
      relative to their own season-long average; $p=0.001$.
      \item Effect size is not considered.
      \item Density of support is questionable.\footnote{What if we miss the only shot we took on Sunday, is that relevant?}     \end{itemize}
\end{frame}

\begin{frame}   \frametitle{Prev. research - likelihood of meeting the spread in NBA}
  Teams under-perform bookmakers expectations on Sunday afternoons when on the road.\footnote{Ezekowitz, ('14).}
\vspace{12pt}
    \begin{itemize}       
    \item Only 129 out of 284 games did the away team meet the spread, 
      using 2006-14 data; $p=0.1089$.
    \item Not quite significant at the 90\% confidence level.     
    \item If we bet \$100 on each Sunday game, we'd only profit \$1,310.\footnote{$(284 - 129) \times 100 + 129 \times (-110) = 1,310$.} 
    \end{itemize}
  \end{frame}

@@

    Conditions on (almost) every conceivable element of the game:
    \begin{itemize}     \item strengths and weaknesses against particular opponents,
    \item momentum,
    \item injuries, and
    \item even possibly a weather forecast.     \end{itemize}

@@

\begin{frame}   \frametitle{Our solution: a quasi-experiment}
  Consider games played back-to-back within 24 hours.

@@

\begin{frame}   \frametitle{NBA data: 2010-11 through 2016-17 season}
  \begin{itemize}     \item Each team plays 82 games per season.
    \item 30 teams in the NBA.
    \item Total of $82 \times 30 / 2 = 1,230$ games per season.  
  \end{itemize}

@@

\begin{frame}   \frametitle{MLB data: 2011 through 2017 season}
  \begin{itemize}     \item Each team plays 162 games per season.
    \item 30 teams in the MLB.
    \item Total of $162 \times 30 / 2 = 2,430$ games per season.   \end{itemize}

@@

\begin{frame}   \frametitle{Bureau of Labor Statistics data: 2010-2016}
  We look toward BLS data which describes the number of business establishments 
  (of a very particular type) in each metropolitan statistical area
  for each quarter of each year.

@@

\begin{frame}
  \frametitle{Possible confounders}
  What might influence game-outcomes apart from player-ability?

@@

\begin{frame}   \frametitle{Feature creation}
  We add features which describe
  \vspace{12pt}  
  \begin{itemize}     
    \item Number of days since the last game (i.e. rest-time),
    \item Lagged changes in possession,\footnote{A crude measure of fatigue incurred from last game.}
    \item Lagged game location, and
    \item Travel Distance.   
  \end{itemize}
  \vspace{12pt}We seek to construct a model which accounts for travel-fatigue and jet-lag.
  For this, we need a little bit more. \end{frame}

\begin{frame}   \frametitle{Accounting for jetlag}
  Prolonged air travel negatively effects sporting performance.\footnote{Lee and Galvez ('12).}

@@

\begin{frame}   
  \frametitle{Travel distance comparison}
  \centering \includegraphics[scale=0.25]{../writing/travel_dist_density_by_party} 
\end{frame}

\begin{frame}   \frametitle{Continuous measure}
  \begin{block}{A-list celebrities tend to party with other celebrities.}     Idea: proxy nightlife using number of musicians in previous location

@@

\begin{frame}   
  \frametitle{Continuous measure of party (NBA)}
  \centering \includegraphics[scale=0.3]{../writing/Party_by_Team_Location} 
\end{frame}

@@

\begin{frame}   \frametitle{Placebo test (discrete indicator): what if players rest?}
  \centering
  \tiny{
\begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{1}{c}{\textit{Dependent variable:}} \\  \cline{2-2}  \\[-1.8ex] & Meet the Spread (NBA) \\  \hline \\[-1.8ex]   Party placebo & 0.072 \\    & (0.082) \\    & \\   Lag changes in posession & 0.00001 \\    & (0.002) \\    & \\   Logged travel distance & $-$0.005 \\    & (0.026) \\    & \\   East-west travel direction & $-$0.099 \\    & (0.231) \\    & \\   Number hours rest time & 0.001$^{*}$ \\    & (0.001) \\    & \\   Time of game during day & 0.001 \\    & (0.013) \\    & \\   Home team effect & $-$0.009 \\    & (0.043) \\    & \\   Logged travel distance * east-west & 0.015 \\    & (0.036) \\    & \\   Constant & $-$0.069 \\    & (0.378) \\    & \\  \hline \\[-1.8ex]  Observations & 9,823 \\  Log Likelihood & $-$6,806.701 \\  Akaike Inf. Crit. & 13,631.400 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}  } \end{frame}

@@

\begin{frame}   
  \frametitle{Continuous measure of party (MLB)}
  \centering \includegraphics[scale=0.3]{../writing/Party_by_Team_Location_mlb}   
\end{frame}

@@

\begin{frame}   \frametitle{Returning to NBA: player performance}
  We additionally look toward actual performance metrics, instead of relying on point-spread data.

@@

\begin{frame}   \frametitle{Future research - NBA}
  Within the NBA
  \begin{itemize}     
    \item Are teams more likely to party after a win or a       loss?
    \item Do teams who aren't going to make the playoffs party more?
    \item Are there demographic differences which influence propensity to party? E.g. are younger, unmarried players more susceptible to our treatment effect?
    \item We could also examine how shots taken vs. shots scored are affected for star players.
  \end{itemize}
  \vspace{12pt}  A preliminary and ad-hoc analysis shows that injury risk increases as a function of party.\footnote{We look for instances where players are put on the bench and stop playing (at least several) games as a function of our nightlife indicator.}
  This might inform coaches which players to bench and when.

@@

\begin{frame}   \frametitle{Future research (MLB and general)}
  \begin{itemize}     
  \item Within the MLB, we can look at statistics such as batting averages or general fielding statistics. 
    \item Possibly prolonged effects; recall our magic 24 hour number.
    \item European soccer leagues.   \end{itemize} \end{frame}

@@

\begin{frame}   \frametitle{Why \# drinking establishments or population won't work}
  \includegraphics[scale=0.3]{../writing/why_drinks_or_population_dont_work} \end{frame}
